D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-16 ootavacuous.tex
Define semantic dependency
Add a chapter defining semantic dependency, proving that the
Fundamental Property holds, and raising some questions.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

>>>>

\section{A Definition of Semantic Dependency}
\label{sec:A Definition of Semantic Dependency}

Semantic dependency is a notoriously difficult concept to define
rigorously and precisely.
A large part of the reason is because it was never a completely clear
concept to begin with, especially when there are multiple accesses to
the variables involved.
In this section we will stick our necks out by offering just such a
definition.
No doubt many people will object to it for various reasons, but we
hope it will provide a starting point for discussions that move the
entire field forward.

The definition given below is applicable only to C++ implementations
that treat all atomic objects as though they are volatile or
quasi-volatile.
For compilers that perform only single-thread analysis, not global
analysis, quasi-volatile is sufficient.
In this setting we can relate abstract and hardware executions by
means of the map of accesses described in
section~\ref{sec:Volatile and Quasi Volatile Accesses}.
The key insight is that this allows us to consider semantic
dependencies at the level of the machine code, where they are much
more tractable.

\subsection{For Compilers Using Single-Thread Analysis}
\label{sec:For Compilers Using Single-Thread Analysis}

In this section we consider implementations whose compilers perform
only single-thread analysis and treat all atomic objects as
quasi-volatile.
This implies that if two different programs contain the same thread
(i.e., the same source code for the functions in the thread), the
machine code generated by the compiler for the thread will be the
same in the two programs.

We begin by recognizing that semantic dependencies are relative to a
particular execution and a particular implementation,
as described in section~\ref{sec:Properties of Semantic Dependencies}.
The same source code may or may not contain a semantic dependency,
according to the details of the execution in question and the
machine code produced by the compiler.
For this reason we will characterize semantic dependencies in a given
abstract execution realized by a given hardware execution.
(While it possible to argue about semantic dependencies in abstract
executions that have no hardware realizations, doing so seems like a
pointless activity.)

Let $A$ be an abstract execution of some program $P$ containing a
thread $T$, and let $H$ be a hardware execution realizing $A$.
Let $W$ be an store in $T$ to an atomic object and let $\{R_i \mid i =
0, \ldots, n\}$ be a set of loads in $T$ from atomic objects on which
$W$ might or might not depend.
We can dispose of one case immediately: If $W$ is omitted in $H$ then
the issue of semantic dependency is moot.
You can give either answer since the question is basically meaningless.
Therefore we'll assume that $W$ is not omitted.
Then:
\begin{quote}
There is a semantic dependency from $\{R_i\}$ to $W$ in $A$ and $H$,
relative to the compiler used to produce $H$, if there is another
abstract execution $B$ realized by hardware execution $G$ under the
same compiler that together \emph{witness the semantic dependency}.
\end{quote}

To be a proper witness, $B$ must be an execution of some program $Q$,
not necessarily the same as $P$ but which contains the same thread $T$.
The thread should start out with the same initial state in $A$ and
$B$, and all loads in $A$ coming before any of the $R_i$ should obtain
the same value as they do in $B$ (this is part of our interpretation
of ``all other things being equal'').
It follows that the two abstract executions of $T$ will be identical
up to the first of the $R_i$ loads.

Let $W'$ and $\{R'_i\}$ be the accesses in $H$ that $W$ and the
non-omitted $\{R_i\}$ loads map to.
We then require that the hardware executions of $T$ in $H$ and $G$ be
identical for an initial period lasting up to the first of the $R'_i$.
Following this initial period there will be a common period, during
which $H$ and $G$ execute the same machine instructions but do not
necessarily compute the same values.
This common period ends when one of the hardware executions
takes a conditional branch that the other doesn't, or when a computed
branch leads to different addresses in the two executions, or when $T$
ends, whichever comes first.
Past this point $H$ and $G$ diverge and are no longer directly
comparable, as they execute different instructions.
Our third requirement for being a witness is that each load in the
common portion either must obtain the same value in $H$ and $G$, or
must itself be one of the $R'_i$ loads, or must be ordered in $H$
after one of the $R'_i$ loads (this is the remaining part of our
interpretation of ``all other things being equal''.)

Finally, we need to determine an instruction $X'$ in $G$ that
corresponds to $W'$.
If $W'$ is part of the initial or common portion of $H$ this is no
problem; we can take $X'$ to be $W'$ itself.
But if $W'$ is in the divergent portion of $H$ then things aren't so
simple.
The choice is somewhat arbitrary, and so we will fall back on the
earlier proposal of matching up stores by the order they occur.
Let \co{y} be the atomic object that $W'$ stores to, and suppose $W'$
is the $N$th store to \co{y} within the divergent portion of $H$.
Then $X'$ will be the $N$th store to \co{y} in the divergent portion
of $G$, if such a store exists.
Our last requirement for being a witness to a semantic dependency
is that $X'$ act differently from $W'$: it doesn't exist, it stores
a different value, or it stores to an object other than \co{y}.

\subsection{For Compilers Using Global Analysis}
\label{sec:For Compilers Using Global Analysis}

As promised earlier, we now consider implementations whose compilers
may use global analysis.
In order to obtain the desired results we have to require that these
compilers treat all atomic objects as volatile.
Equivalently, the machine code generated by such a compiler must be
the same for a given program as for a ``volatilized'' form of the
program in which all atomic objects are defined to be volatile.

In this context our definition of semantic dependency is essentially
the same as before.
Since we can no longer expect the machine code for a thread to be the
same regardless of the program it belongs to, the program $Q$ in the
earlier definition (of which $B$ and $G$ are executions) must now be
$P$ or its volatilized form.
However, we do now allow the possibility that the executions $B$ and
$G$ take place in a non-benign environment.
Aside from these minor adjustments, the definition remains unchanged.

\subsection{Verifying the Fundamental Property}
\label{sec:Verifying the Fundamental Property}

Of course we want to check that our definition of semantic dependency
satisfies the Fundamental Property of
section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
Given the information we have already presented, the demonstration is
easy.

Suppose we have $W$, $\{R_i\}$, $A$, and $H$ as in the definition.
The Fundamental Property assumes that $W$ is not omitted in $H$, so
there is an abtract execution $B$ with hardware realization $G$
witnessing the semantic dependency of the store $W$ on the loads
$\{R_i\}$ in $A$ and $H$.
We must show that some $R_i$ is not omitted and $R'_i$ is ordered
before $W'$ in $H$.
The proof splits into three cases.

First case: $W'$ lies in the initial portion of $H$.
During the initial portion of the hardware executions, $H$ and $G$
behave identically and therefore $W'$ performs the same write in
both.
This contradicts the fact that $B$ and $G$ witness the semantic
dependency.

Second case: $W'$ lies in the common portion of $H$.
Since the action of $W'$ in $H$ is different from its action in $G$,
at least one of its inputs must differ between the two hardware
executions.
Therefore the source instruction for that input must behave
differently, and so must one of its sources, going back until we reach
a load instruction that obtains differing values in $H$ and $G$.
Then $W'$ depends on this load and so is ordered after it.
And since the load must lie in the common portion of $H$, by the
definition of semantic dependency it must either be one of the
$R'_i$ or be ordered after one of them.
Therefore $W'$ is ordered after one of the $R'_i$ in $H$,
which certainly means that $R_i$ is not omitted.

Third case: $W'$ lies in the divergent portion of $H$.
This happens when $W'$ comes after the conditional or computed branch
which marks the end of the common period by going different ways in
$H$ and $G$.
Just as in the previous case, since the branch behaves differently in
the two executions it must be ordered after one of the $R'_i$ loads.
And then so must $W'$, because any instruction following a conditional
or computed branch instruction is ordered after the branch.
QED.

A corollary of this result is that if an implementation's compiler
either
\begin{itemize}
\item	uses single-thread analysis and treats atomic objects as
	quasi volatile, or
\item	uses global analysis and treats atomic objects as volatile,
\end{itemize}
then programs produced by that implementation will never exhibit OOTA.
Thus the implementation will automatically comply with full C++, even
though it may be been designed only to comply with loose C++.

\subsection{Outstanding Issues}
\label{sec:Outstanding Issues}

Here we consider some general questions related to our definition of
semantic dependency.

\subsubsection{Relative versus Absolute Dependency}
\label{sec:Relative versus Absolute Dependency}

A drawback of the definition is that it is only relative to a specific
compiler or implementation.
This may strike some people as too compromising, in that a given piece
of code should either contain or not contain a semantic dependency
independent of the implementation used to run it.

We can address this drawback by defining an \emph{absolute semantic
dependency} as one that is present relative to any valid C++
implementation, past, present, or future.
Of course this notion has its own problems, including that it is
extremely non-constructive and impossible to apply in practice.
However it may be the best we can do with our current understanding
of computing systems.

There is one thing we can definitely state:
Programs produced by an implementation of the right sort will never
exhibit absolute OOTA (that is, an OOTA cycle in which all the
semantic dependencies are absolute).
This is simply because a semantic dependency that is absolute is {\it
a fortiori\/} a semantic dependency relative to the compiler in use.

But in fact we have shown more than this:
Programs will never exhibit an OOTA cycle relative to the compiler
used to build them, even when that cycle is not absolute.
In this sense we have gone beyond the requirement of full C++.

\subsubsection{Global Analysis and Volatile versus Quasi Volatile}
\label{sec:Global Analysis and Volatile versus Quasi Volatile}

In principle we don't need to require global-analysis compilers to
treat atomic objects as volatile; our results would still hold if they
merely treated them as quasi volatile.
We chose not to do this because it would violate our intuitions about
semantic dependencies.

For example, consider the Simple OOTA program, repeated here in
simplified form:
\begin{quote}
\begin{verbatim}
void thread1() {
   int r1 = x;
   y = r1;
}

void thread2() {
   int r2 = y;
   x = r2;
}
\end{verbatim}
\end{quote}
A compiler using global analysis and treating \co{x} and \co{y} as
quasi-volatile objects could omit the two loads, replacing them in the
machine code with simple assignments ``\co{r1 = 42}'' and
``\co{r2 = 42}''.
This would be a valid transformation, and the resulting behavior would
not count as OOTA because according to our definition, the
dependencies in \co{thread1} and \co{thread2} would not be semantic.

To see why, recall that a semantic dependency must have a witness,
another execution in which the store acts differently.
But this transformed program has no other executions; every time it
runs it will write 42 to both \co{x} and \co{y}.
(Keep in mind also that since the atomic objects are not treated as
volatile, they are not subject to unspecified interference when the
program runs in a non-benign environment.)

This unintuitive behavior could not occur if the two loads were not
omitted.
In fact, the definition of semantic dependency might remain
perfectly acceptable if the requirement for global-analysis compilers
were weakened, if the compiler treated atomic objects as quasi
volatile and in addition was not allowed to omit accesses to them.
This is a possible topic for future research.

\subsubsection{Effect of Memory Layout}
\label{sec:Effect of Memory Layout}

Part of our demonstration of the Fundamental Property of semantic
dependencies relies on the fact, stated in
section~\ref{sec:Abstract Executions},
that an abstract execution of a thread is entirely determined by the
values obtained for its loads.
But when we compare abstract executions of the same thread in two
different programs, this may no longer be entirely true owing to the
effect of differing memory layouts.

Consider this simple example:
\begin{quote}
\begin{verbatim}
x = (int) &x;
\end{verbatim}
\end{quote}
Even though the example contains no loads at all, it may store
different values when running in different programs because the
variable \co{x} may be allocated at differing addresses in those
programs.
According to our definition, this could count as a degenerate OOTA
cycle of length one, in which the store is semantically dependent on
an empty set of loads!

To rule out such pathological counterexamples we should require that
in a witness to a semantic dependency, the addresses of all the
objects and functions referred to in the thread $T$ are the same as
in the original execution.
This is a very technical restriction but there are occasions when
the issue might realistically arise, such as when computing a hash
value based on an object's address.

\subsubsection{Merging Quasi-Volatile Loads}
\label{sec:Merging Quasi-Volatile Loads}

The compiler is permitted to merge quasi-volatile loads.
This can lead to surprising results because a particular load
may be merged with an earlier load in one execution and with a
later load in another.
This is demonstrated in the following, which is a variant of
the example in
section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}:
\begin{quote}
\begin{verbatim}
int r1 = (x != 0);
int r2 = (x != 0);
int r3 = (x != 0);
int r4 = y;
z = (r2 == r4);
\end{verbatim}
\end{quote}
Consider an abstract execution in which \co{r1}, \co{r2}, and \co{r4}
are zero and \co{r3} is one (because another thread changed the value
of \co{x} between two of the loads).
We would expect that the store to \co{z} would be semantically
dependent on the load of \co{y} in this execution.

However, a single-thread analysis compiler can translate this into the
machine code equivalent of:
\begin{quote}
\begin{verbatim}
int r1 = (x != 0);
int r2;
int r3 = (x != 0);
int r4 = y;
if (r1 != r3) {
   r2 = r4;
   z = 1;
} else {
   r2 = r1;
   z = (r2 == r4);
}
\end{verbatim}
\end{quote}
In effect, the \co{r2} load is merged with the \co{r1} load in
executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
merged with the \co{r3} load in other executions.

To demonstrate the semantic dependency, a suitable witness would have
to include a hardware realization of the abstract execution in which
\co{r1} and \co{r2} are zero and \co{r3} and \co{r4} are one.
But there are no hardware realizations of this execution with the
machine code indicated above!
Since \co{r1} is different from both \co{r3} and \co{r4}, the \co{r2}
load will be merged with the \co{r3} load and so \co{r2} will
necessarily be one, not zero.
Thus, relative to this compiler the example does not contain a
semantic dependency.

Although it is unexpected, we cannot say that this conclusion is
definitely wrong, because our intuitive notions of semantic
dependency are not clear in cases where multiple loads of the same
variable are present.

\section{To Be Moved Or Removed...}

@@@ Review
